SNP Analysis

### 2019-12-09 snp/SNP_Prediction_Analysis.Rmd
cleaning things up

@@

model.dir = '~/Data/Ellie/Analysis/models_5e6_all/50kb'
load(paste0(model.dir,'/model_data.Rdata'), verbose = F)
load(paste0(model.dir,'/all.pt.alpha.Rdata'), verbose = F)

alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]

@@

patient.info = readxl::read_xlsx('~/Data/BarrettsProgressionRisk/Analysis/SNP/metadata_T1T2.xlsx') %>% 
  mutate( UniqueSampleID = paste(PatientID, `Timepoint Code`, sep='_'), Path.Status = Status, PatientID = as.character(PatientID)) %>% 
  mutate(Path.Status = as.character(ifelse(Pathology %in% c('IMC','HGD'), 'P', Path.Status)))

@@

@@

qcdata = qcdata %>% mutate(
  ASCAT.SCA.ratio = purrr::pmap_dbl(list(PatientID, Samplename, Ploidy), .f=calc.ratio),
  Level = toupper(Level),
  SampleType = case_when(
    grepl('BLD', Level, ignore.case = T) ~ 'Blood Normal',
    grepl('gastric', Level, ignore.case = T) ~ 'Gastric Normal',
    TRUE ~ 'BE'
))

@@

sample.info = left_join(sample.info, patient.info[,c('PatientID','Timepoint','Timepoint Code','Status','Pathology')], by=c('PatientID','EndoID' = 'Timepoint Code')) %>%
  mutate(Status = factor(as.character( ifelse(is.na(Status) & Level %in% c('BLD','GASTRIC'), 'Normal', Status))))
head(sample.info)

@@

@@

ASCAT and Xiahong's estimates show a `r round(ct$estimate,2)` correlation. That suggests that the remainder is due to the LOH I'm not trying to call.

@@

There's a significant difference in the numbers of NDBE or HGD samples between the groups
```{r, echo=F}
pander(chisq.test(table(patient.info$Pathology, patient.info$Status)['HGD',]), caption='HGD')
pander(chisq.test(table(patient.info$Pathology, patient.info$Status)['NDBE',]), caption='NDBE')

# Most progressors are HGD at timepoint 1
pander( round(with(subset(sample.info, Endoscopy == 1), table(Pathology, Status))['HGD',]/table(subset(sample.info, Endoscopy == 1)$Status), 3)*100, caption='% HGD at T1')
```

@@

```{r, echo=F, warning=F, message=F}
qcdata = left_join(qcdata, sample.info[,c('Samplename','Status','Timepoint','SCA.Ratio')], by='Samplename')

@@

low = subset(qcdata, SampleType == 'BE' & ASCAT.SCA.ratio <= median(subset(qcdata, SampleType != 'BE')$ASCAT.SCA.ratio))

anno = cbind.data.frame(table(subset(qcdata, SampleType == 'BE' & ASCAT.SCA.ratio <= median(subset(qcdata, SampleType != 'BE')$ASCAT.SCA.ratio))$Status)/table(subset(qcdata, SampleType == 'BE')$Status))
colnames(anno) = c('Status','Ratio')
anno$label='median(normal SCA)'
anno$x = 0.4
anno$y = median(subset(qcdata, SampleType != 'BE')$ASCAT.SCA.ratio)

ggplot(subset(qcdata,SampleType == 'BE'), aes(Purity, ASCAT.SCA.ratio, color=Status, group=Status)) + geom_point() + facet_grid(~Status, scales = 'free') + ylim(0,0.01) + 
  geom_hline(yintercept = anno$y[1]) + geom_text(data=anno, aes(x, y, label=label), nudge_y = 0.0001) +
  geom_text(data=anno, aes(x, y, label=paste(round(Ratio,2)*100,'% samples',sep='')), nudge_y = -0.0001)

ggplot(qcdata, aes(Ploidy, ASCAT.SCA.ratio, color=Status)) + geom_point() + facet_grid(~SampleType)

ggplot(qcdata, aes(Ploidy, ASCAT.SCA.ratio, color=Status)) + geom_point() + facet_grid(~Status)

ggplot(subset(qcdata,SampleType == 'BE'), aes(Timepoint, ASCAT.SCA.ratio, color=PatientID, group=Timepoint)) + geom_boxplot() + 
  facet_wrap(~Status, scales='free_x', ncol=2) + labs(title='SCA Ratio by Timepoint') + theme(legend.position = 'none')

ggplot(qcdata, aes(round(Ploidy,2))) + geom_histogram() + 
  facet_wrap(~SampleType, ncol=1) + labs(x='Ploidy', title='Ploidy, normal vs BE')

@@

#x=median(qcdata$SCA.Ratio,na.rm=T)+sd(qcdata$SCA.Ratio,na.rm=T)
x=0.05
y=median(qcdata$Purity,na.rm=T)-sd(qcdata$Purity,na.rm=T)/2

ggplot(qcdata, aes(Purity, ASCAT.SCA.ratio, color=ASCAT.SCA.ratio < 0.2)) + geom_point() + 
         facet_wrap(~Status, ncol=1) + labs(title='Purity~SCA.Ratio') + geom_vline(xintercept=y, color='grey')

ggplot(qcdata, aes(ASCAT.SCA.ratio, Purity, color=Status)) + geom_point() + 
  facet_wrap(~Status) + labs(x='SCA', title='SCA vs Purity') + 
  geom_vline(xintercept = x, color='grey') + geom_hline(yintercept = y, color='grey')

table(unique(subset(qcdata, Purity > y & ASCAT.SCA.ratio < x & SampleType == 'BE', select=c('Status','PatientID')))$Status)

ggplot(qcdata, aes(Purity)) + geom_histogram() + geom_vline(xintercept = y, col='red') + 
  facet_wrap(~SampleType, scales = 'free') + labs(title='Purity by sampletype')

@@

ggplot(subset(qcdata, SampleType == 'BE'), aes(round(Ploidy,2), Purity, color=Status)) + geom_point() + facet_grid(~Status) + 
  labs(title='BE only, purity vs ploidy') + geom_hline(yintercept=y, color='grey')

ggplot(qcdata, aes(ASCAT.SCA.ratio, color=Status, fill=Status)) + facet_wrap(~Status, scales='free_y',nrow=1) +  geom_density(alpha=0.5) + labs(x='SCA Ratio', title='SCA Ratio, NP vs P')

ggplot(qcdata, aes(ASCAT.SCA.ratio, color=Status, fill=Status)) + facet_wrap(Status~SampleType, scales='free_y',nrow=2) +  geom_density(alpha=0.5) + labs(title='SCA Ratio, NP vs P')

summary(subset(qcdata, SampleType == 'BE')$ASCAT.SCA.ratio)
summary(subset(qcdata, SampleType != 'BE')$ASCAT.SCA.ratio)

ggplot(subset(qcdata, SampleType == 'BE'), aes(Purity, ASCAT.SCA.ratio, color=Status)) + facet_grid(~Status) +
  geom_point() + geom_vline(xintercept = y, color='darkgreen') + geom_hline(yintercept = x, color='red') + 
  labs(title='Barretts only')

ggplot(subset(qcdata, SampleType == 'BE'), aes(Ploidy, ASCAT.SCA.ratio, color=Status)) + facet_grid(~Status) + geom_point() + 
  geom_hline(yintercept = 0.02, color='red') + labs(title='Barretts only')

## PER ENDOSCOPY
qcdata.endo = qcdata %>% group_by(PatientID, EndoID, Status, Timepoint, SampleType) %>% dplyr::summarise_if(is.numeric, c('mean','max','min','sd')) %>% mutate_if(is.numeric, round, digits=3)
qcdata %>% dplyr::group_by( Status, SampleType) %>% dplyr::summarise( median(ASCAT.SCA.ratio), sd(ASCAT.SCA.ratio) )

#ggplot(subset(qcdata.endo, SampleType == 'BE'), aes(ASCAT.SCA.ratio_max, Purity_max)) + geom_point() + facet_grid(~Status) + geom_hline(yintercept = 0.95)  + geom_vline(xintercept = 0.001)

lowsca = subset(qcdata.endo, SampleType == 'BE' & ASCAT.SCA.ratio_max <= 0.01 & Purity_max > 0.98 )

#table(unique(lowsca[,c('PatientID','Status')])$Status)/table(unique(qcdata.endo[,c('PatientID','Status')])$Status)

#table(unique(subset(qcdata.endo, SampleType == 'BE' & ASCAT.SCA.ratio_max <= 0.001, select=c('PatientID','Status')))$Status)

ggplot(subset(qcdata.endo,SampleType == 'BE'), aes(Status, ASCAT.SCA.ratio_max, group=Status)) + geom_boxplot() + geom_hline(yintercept = 0.002, color='red')

qcdata %>% dplyr::group_by( Status, SampleType) %>% dplyr::summarise( median(ASCAT.SCA.ratio))

@@

CN Adjustment uses Annalise's whole-genome non-progressor samples to generate a set of mean and SD values per copy-number state. The median LogR values are adjusted using the mean and SD based on the CN state.  The adjusted LogR values are then winsorize per sample, and tiled in 5Mb bins.

```{r, warning=F, echo=F, message=F, fig.height=10, fig.width=8}

if (file.exists('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata')) {
  load('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata', verbose=T) 
} else {
  ptdirs = list.dirs('~/Data/Reid_SNP/PerPatient', recursive = F)
  mergedSegs = NULL;  mergedArms = NULL
  length(ptdirs)

@@

# Adjust for ploidy -- this was done in the processing
# for (i in 1:nrow(mergedSegs)) {
#   sample = rownames(mergedSegs)[i]
#   x = unlist(strsplit(sample, '_'))
#   #ploidy = mean(subset(qcdata, SampleType == 'BE' & PatientID == x[1] & EndoID == x[2])$Ploidy)
#   ploidy = subset(qcdata.endo, PatientID == x[1] & EndoID == x[2] & SampleType == 'BE')$Ploidy_mean
#   if (grepl('BLD|gastric', sample, ignore.case = T)) ploidy = 2
#   mergedSegs[i,] = mergedSegs[i,]-(ploidy-1)
#   mergedArms[i,] = mergedArms[i,]-(ploidy-1)
# }

@@

nm.rows = grep('BLD|GASTRIC', rownames(mergedSegs))
#normalSegs = mergedSegs[nm.rows,]
#normalArms = mergedArms[nm.rows,]

# nrow(normalSegs) == nrow(normalArms)
# 
# mergedSegs = mergedSegs[-nm.rows,]
# mergedArms = mergedArms[-nm.rows,]
# 
# nrow(mergedSegs) == nrow(mergedArms)

lowSeg = mergedSegs[grep('^403', rownames(mergedSegs), value=T),,drop=F]
p1 = ggplot(melt(lowSeg), aes(Var2, value)) + facet_grid(~Var1, space='free_x', scales='free') + geom_point() +
  theme(axis.text.x=element_blank() ) + labs(title='Low SCA samples', subtitle='patient 403')

highSeg = mergedSegs[grep('^512', rownames(mergedSegs), value=T),,drop=F]
p2 = ggplot(melt(highSeg), aes(Var2, value)) + facet_grid(~Var1, space='free_x', scales='free') + geom_point() +
  theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA samples', subtitle='patient 512')

grid.arrange(p1,p2,ncol=1)
```

# Values in the binned segments

Blood/gastric normals have very little variance compared to all of the BE samples (no idea which is P vs NP)

```{r normals, message=F, fig.height=10, fig.width=8}
#nm = melt(normalSegs)
nm = melt(mergedSegs[nm.rows,])
be = melt(mergedSegs[-nm.rows,])

lims = range(nm$value, be$value, na.rm=T)

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric',subtitle=paste('variance=',var(nm$value),sep=''),x='') + 
    theme(axis.text.x = element_blank()) + ylim(lims),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples',subtitle=paste('variance=',var(be$value,na.rm=T),sep=''),x='') + 
    theme(axis.text.x = element_blank()) + ylim(lims)
)

@@

   ggplot(melt(t(adjust.cols(mergedSegs, z.mean, z.sd)[randNormal,])), aes(x=Var2, y=value)) + geom_point() + 
    theme(axis.text.x=element_blank()) + labs(title='Col adjusted CN, binned'),
 top='Random BLD or gastric normal', bottom=randNormal)

@@

   ggplot(melt(t(adjust.cols(mergedSegs, z.mean, z.sd)[randBE,])), aes(x=Var2, y=value)) + geom_point() + 
    theme(axis.text.x=element_blank()) + labs(title='Col adjusted CN, binned'),
top='BE', bottom=randBE)

@@

Adjust using the RAW values...

```{r raw.adjust, fig.height=16, fig.width=4, echo=T}
q.norm.by<-function(df, target) {
  dfq = do.call(cbind.data.frame,lapply(1:ncol(df), function(i) {
    as.data.frame(normalize.quantiles.use.target(df[,i,drop=F], as.vector(target[,i])))
  }))

  colnames(dfq) = colnames(df)
  rownames(dfq) = rownames(df)

  return(as.matrix(dfq))
}

#copySegs = q.norm.by(mergedSegs, raw.segs)
#copyArms = q.norm.by(mergedArms, raw.arms)

training = do.call(bind_rows, purrr::map(list.files('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb', '5e06_tiled_segvals.txt', full.names = T, recursive = T), function(f) {
  read_tsv(f, col_types = cols(.default=col_double(), 'X1'=col_character()))
}))
training.segs = as.matrix(training[,-1]); rownames(training.segs) = training$X1

training = do.call(bind_rows, purrr::map(list.files('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb', 'arms_tiled_segvals.txt', full.names = T, recursive = T), function(f) {
  read_tsv(f, col_types = cols(.default=col_double(), 'X1'=col_character()))
}))
training.arms = as.matrix(training[,-1]); rownames(training.arms) = training$X1

median(training.segs)
median(mergedSegs)
median(mergedSegs + 0.98)

median(training.arms)
median(mergedArms)
median(mergedArms + 0.98)

mergedSegs = mergedSegs + 0.98
mergedArms = mergedArms + 0.98

merged = rbind(mergedSegs, training.segs)
merged.mean = apply(merged, 2, mean, na.rm=T)
merged.sd = apply(merged, 2, sd, na.rm=T)

@@

# grid.arrange(
#   ggplot( melt(raw.segs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, sWGS bins (no arms)'),
#   ggplot( melt(mergedSegs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP bins (no arms)')
#   ggplot( melt(copySegs), aes(sample=value)) + stat_qq() + labs(title='Normal Q-Q plot, SNP bins q-normed')
# )

normal.ids = sample.info %>% filter(grepl('BLD|gastric', Level, ignore.case=T)) %>% dplyr::summarise(rows = paste0(PatientID, '_', EndoID)) %>% pull

nm.rows = which(rownames(mergedSegs) %in% normal.ids)

@@

```{r echo=F,message=F, fig.width=10, fig.height=8}
#cxN = score.cx(ns, 1)
#arrayDfNorm = subtract.arms(ns, na)
#arrayDfNorm = cbind(ns, na)
#arrayDfNorm = cbind(arrayDfNorm, 'cx'=unit.var(cxN, mn.cx, sd.cx))

#probN = predict(fitV, newx=arrayDfNorm, s=l, type='response')
#length(which(probN > 0.5))/nrow(probN) # FPR
#length(which(probN <= 0.5))/nrow(probN)

#preds.normal = cbind.data.frame(rownames(probN),probN, predict(fitV, newx=arrayDfNorm, s=l, type='link'))
#colnames(preds.normal) = c('Samplename','Prob', 'RR')

cx = BarrettsProgressionRisk:::scoreCX(mergedSegs, 1)
#arrayDf = cbind(ms,ma)
arrayDf = BarrettsProgressionRisk:::subtractArms(mergedSegs, mergedArms)
#arrayDf = cbind(arrayDf, 'cx'=unit.var(cx, mn.cx, sd.cx))
arrayDf = cbind(arrayDf, 'cx'=BarrettsProgressionRisk:::unit.var(cx, mn.cx, sd.cx))

lims=range(arrayDf, dysplasia.df,na.rm=T)
grid.arrange(
  ggplot(melt(arrayDf), aes(Var2, value)) + geom_point() + ylim(lims) +
    geom_point(data=melt(arrayDf[,589:634]), col='red') + labs(title='SNP'),
  ggplot(melt(dysplasia.df), aes(Var2, value)) + geom_point() + ylim(lims) +
      geom_point(data=melt(dysplasia.df[,589:634]), col='red') + labs(title='sWGS'),
nrow=1)

@@

```{r, warning=F, fig.height=8}

prob = predict(fitV, newx=arrayDf, s=l, type='response')
rr = predict(fitV, newx=arrayDf, s=l, type='link')

preds = bind_cols('Samplename'=rownames(prob),'Prob'=prob[,1],'RR'=rr[,1]) %>%
  mutate(Risk = case_when(
    Prob <= 0.3 ~ 'Low',
    Prob > 0.3 & Prob < 0.5 ~ 'Moderate',
    Prob >= 0.5 ~ 'High' 
))

normals = filter(preds, grepl('BLD|gastric', Samplename, ignore.case=T))
table(normals$Risk)

@@

preds = preds %>% left_join(patient.info %>% dplyr::select(UniqueSampleID, Pathology, Status), by=c('Samplename'='UniqueSampleID')) %>%
   mutate(Status = as.character(ifelse(Pathology %in% c('HGD','IMD'), 'P', Status)))

@@

hist(preds$Prob)

length(which(normals$Prob > 0.5))/nrow(normals) # FPR
length(which(normals$Prob < 0.5))/nrow(normals)

myPal = rev(RColorBrewer::brewer.pal(11, 'RdYlBu'))
grid.arrange(
  ggplot(preds, aes(RR)) + geom_histogram(aes(fill=..x..), bins=15, show.legend = F) +
    scale_fill_gradientn(colors = myPal,  name='') + 
    labs(y='n Samples', x='Relative Risk', title='Unadjusted relative risk'),
  # ggplot(preds, aes(`Adj. RR`)) + geom_histogram(aes(fill=..x..), bins=15, show.legend = F) +
  #   scale_fill_gradientn(colors = myPal,  name='') +
  #   labs(y='n Samples', x='Relative Risk', title='Adjusted relative risk'),
top='Relative risk')

>>>>

dir.create('plots/snp',recursive = T, showWarnings = F)

ggplot2::theme_set(theme_minimal())

@@

model.dir = '~/Data/BarrettsProgressionRisk/Analysis/models_5e6_all/50kb'
load(paste0(model.dir,'/model_data.Rdata'), verbose = T)
load(paste0(model.dir,'/all.pt.alpha.Rdata'), verbose = F)

train.labels = labels
alpha = '0.9'
fitV = models[[alpha]]
l = performance.at.1se[[alpha]]$lambda
features = coefs[[alpha]]

@@

patient.info = readxl::read_xlsx('~/Data/BarrettsProgressionRisk/Analysis/SNP/metadata_T1T2.xlsx') %>% 
  mutate( UniqueSampleID = paste(PatientID, `Timepoint Code`, sep='_'), Path.Status = Status, PatientID = as.character(PatientID))
#%>%  group_by(PatientID) %>% mutate(Path.Status = as.character(ifelse(length(which(grepl('IMC|HGD',Pathology))) > 0, 'P', Path.Status)))

path.np.ids = patient.info %>% filter(Pathology %in% c('HGD', 'IMC') & Status == 'NP') %>% dplyr::select(PatientID) %>% distinct %>% pull
patient.info = patient.info %>% mutate(Path.Status = ifelse(PatientID %in% path.np.ids, 'P', Path.Status))

@@

# swgs.seg.count = read_tsv('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb/segment_counts.tsv', col_types = 'dc')
# 
# # All segments
# seg.counts = do.call(bind_rows, purrr::map(segments.list, function(sl) {
#   sl %>% dplyr::mutate(Total = nMajor+nMinor) %>% group_by(sample, Total) %>% 
#     tally %>% #filter(Total != 2) %>% 
#     ungroup %>% group_by(sample) %>% dplyr::summarise(n = sum(n))
# }))

#plot(seg.counts$n)
#points(swgs.seg.count$`n segs`, col='red')

@@

qcdata = qcdata %>% mutate(
  ASCAT.SCA.ratio = purrr::pmap_dbl(list(PatientID, Samplename, Ploidy), .f=calc.ratio),
  Level = toupper(Level),
  SampleType = case_when(
    grepl('BLD|gastric', Level, ignore.case = T) ~ 'Blood/Gastric Normal',
    TRUE ~ 'BE'
))

@@

sample.info = left_join(sample.info, dplyr::select(patient.info,c('PatientID','Timepoint','Timepoint Code','Status','Pathology')), by=c('PatientID','EndoID' = 'Timepoint Code')) %>%
  mutate(Status = factor(as.character( ifelse(is.na(Status) & Level %in% c('BLD','GASTRIC'), 'Normal', Status))))
head(sample.info)

@@

#seg.counts %>% dplyr::summarise_at(vars(-`2`), )
#seg.counts = left_join(seg.counts, dplyr::select(qcdata,c(Samplename, Ploidy, Purity, ASCAT.SCA.ratio, SampleType)), by=c('sample'='Samplename'))
# seg.counts %>% filter(ASCAT.SCA.ratio < 0.01 & Purity > 0.98)
# ggplot(filter(seg.counts, !is.na(Ploidy)), aes(n,ASCAT.SCA.ratio)) + facet_grid(~SampleType) + ylim(0,0.05) + xlim(0,200) +
#  geom_point(aes(color = ASCAT.SCA.ratio < 0.01 & Purity > 0.98)) + theme(legend.position = 'none')

@@

# Brief processing notes

ASCAT was run on each SNP sample to obtain SCA calls. ASCAT and Xiahong's estimates show a `r round(ct$estimate,2)` correlation.

Prior to predicting the samples, in the step where I create 5Mb bins across the genome I merge the samples per endoscopy due to differences in the sampling. The sWGS had multiple biopsies aggregated per-level, while the SNP data was a single 'research' biopsy per level.

@@

There's a significant difference in the numbers of NDBE or HGD samples between the groups at T1.
```{r, echo=F}
#pander(chisq.test(table(patient.info$Pathology, patient.info$Status)['HGD',]), caption='HGD')
#pander(chisq.test(table(patient.info$Pathology, patient.info$Status)['NDBE',]), caption='NDBE')

# Most progressors are HGD at timepoint 1
sample.info %>% filter(Timepoint == 'T1') %>% group_by(Status, Pathology) %>% tally %>% spread(Pathology,n) %>% kable(caption='Number of samples by pathology at T1') %>% kable_styling(full_width = F)

@@

`r qcdata %>% group_by(SampleType) %>% tally %>% spread(SampleType,n) %>% kable(caption='Sample numbers') %>% kable_styling(full_width=F)`

```{r, echo=F, warning=F, message=F}

qcdata = left_join(qcdata, dplyr::select(sample.info,c('Samplename','Status','Timepoint','SCA.Ratio')), by='Samplename')

@@

low = filter(qcdata, SampleType == 'BE' & ASCAT.SCA.ratio <= median(filter(qcdata, SampleType != 'BE')$ASCAT.SCA.ratio))

low %>% group_by(Status, Timepoint) %>% tally %>% spread(Timepoint,n) %>% kable(caption='BE Samples with an ASCAT SCA ratio below the median ratio of the blood/gastric normal samples') %>%
  kable_styling(full_width = F)

qcdata = qcdata %>% mutate(SampleType = case_when(grepl('Normal',SampleType) ~ 'Blood/Gastric Normal', TRUE ~ 'BE'),)

anno = qcdata %>% group_by(SampleType) %>% dplyr::summarise(Ratio = median(ASCAT.SCA.ratio) ) %>% rowwise %>% 
  mutate(label = paste0('median(',SampleType,')','=',round(Ratio,3)))

ggplot(qcdata, aes(Status, ASCAT.SCA.ratio)) + ylim(0,0.05) + 
  facet_grid(~SampleType, scales='free_x') +
  geom_jitter(width=0.1, aes(color=Status)) + 
  geom_hline(data=dplyr::rename(anno, Status=SampleType), aes(yintercept = Ratio, linetype=Status), color='grey42') +
  geom_label(data=dplyr::rename(anno, Status=SampleType), aes(x='NP', y=Ratio, label=label), alpha=0.5)

ggplot(filter(qcdata, !is.na(Ploidy)), aes(factor(round(Ploidy)), ASCAT.SCA.ratio, color=Status)) + 
  facet_grid(~SampleType, scales = 'free_x', space='free') + geom_boxplot() + labs(x='Ploidy')

ggplot(filter(qcdata,SampleType == 'BE'), aes(Timepoint, ASCAT.SCA.ratio, fill=Timepoint, group=Timepoint)) + geom_jitter(width=0.2) +
  geom_boxplot(alpha=0.5) + facet_wrap(~Status, scales='free_x', ncol=2) + labs(title='SCA Ratio by Timepoint') + theme(legend.position = 'none')

@@

x = 0.01
y = round(mean(filter(qcdata, SampleType != 'BE')$Purity) - sd(filter(qcdata, SampleType != 'BE')$Purity),2)

p = ggplot(qcdata, aes(Purity, ASCAT.SCA.ratio)) + facet_wrap(~SampleType, nrow=1) + 
  geom_point(aes(color=ASCAT.SCA.ratio < x | Purity > y), show.legend = F) + 
  geom_vline(xintercept=y, color='grey') + 
  geom_hline(yintercept=x, color='grey') +
  labs(title='Purity~SCA.Ratio') + theme(text = element_text(size=12))
p
ggsave('plots/snp/purity_v_sca.png', plot=p, height=4, width=4, units = 'in', dpi=300)

@@

ggplot(subset(qcdata, SampleType == 'BE'), aes(Ploidy, ASCAT.SCA.ratio, color=Status)) + facet_grid(~Status) + geom_point() + 
  geom_hline(yintercept = x, color='red') + labs(title='Barretts only')

## PER ENDOSCOPY
qcdata.endo = qcdata %>% group_by(PatientID, EndoID, Status, Timepoint, SampleType) %>% 
  dplyr::summarise_if(is.numeric, c('mean','max','min','sd')) %>% mutate_if(is.numeric, round, digits=3)

#qcdata %>% dplyr::group_by( Status, SampleType) %>% dplyr::summarise( median(ASCAT.SCA.ratio), sd(ASCAT.SCA.ratio) )
# lowsca = filter(qcdata.endo, SampleType == 'BE' & ASCAT.SCA.ratio_mean < 0.01 & Purity_max > 0.98 )
# nrow(lowsca)
# nrow(filter(qcdata.endo, SampleType == 'BE' & ASCAT.SCA.ratio_mean >= 0.01 & Purity_max <= 0.98))

tmp = qcdata.endo %>% filter(!is.na(Ploidy_mean)) %>% mutate(Ploidy_mean = factor(round(Ploidy_mean)))
p = ggplot(tmp, aes(Ploidy_mean, ASCAT.SCA.ratio_mean, group=Ploidy_mean)) + 
  facet_grid(~SampleType) +
  geom_jitter(width=0.2, aes(color=Ploidy_mean)) + 
  geom_boxplot(outlier.colour = NA, alpha=0.5) + 
  geom_hline(yintercept = 0.01, color='grey39') + labs(title='Merged per endoscopy', x='Mean ploidy') + 
  theme(text=element_text(size=12), legend.position = 'none')
ggsave('plots/snp/sca_v_ploidy.png', plot=p, height=4, width=4, units = 'in', dpi=300)

@@

I use the raw ASCAT total CN values, adjusted for the sample ploidy, to create the weighted average values in each 5Mb and arm bin of the genome. Note again that at this point the tiled values are per *endoscopy* not per sample.

```{r, warning=F, echo=F, message=F, fig.height=10, fig.width=8}
# Adjust for ploidy -- this was done in the processing script
if (file.exists('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata')) {
  load('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata', verbose=T) 
} else {
  ptdirs = list.dirs('~/Data/Reid_SNP/PerPatient', recursive = F)
  mergedSegs = NULL;  mergedArms = NULL
  length(ptdirs)

@@

@@

nm.rows = grep('BLD|GASTRIC', rownames(mergedSegs))

lowSeg = mergedSegs[grep('^1001', rownames(mergedSegs), value=T),,drop=F]
p1 = ggplot(melt(lowSeg), aes(Var2, value)) + ylim(-1,1) + facet_grid(~Var1) + geom_point() + 
  theme(axis.text.x=element_blank() ) + labs(title='Low SCA samples', subtitle='patient 1001 NP', y='5Mb bins')

lowSeg = mergedSegs[grep('^403', rownames(mergedSegs), value=T),,drop=F]
p2 = ggplot(melt(lowSeg), aes(Var2, value)) +  ylim(-1,1) + facet_grid(~Var1) + geom_point() + 
  theme(axis.text.x=element_blank() ) + labs(title='Low SCA samples', subtitle='patient 403 P', y='5Mb bins')

grid.arrange(p1,p2,ncol=1)

highSeg = mergedSegs[grep('^1114', rownames(mergedSegs), value=T),,drop=F]
p3 = ggplot(melt(highSeg), aes(Var2, value)) +  ylim(-1,1) + facet_grid(~Var1) + geom_point() + 
  theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA samples', subtitle='patient 1114 NP', y='5Mb bins')

highSeg = mergedSegs[grep('^512', rownames(mergedSegs), value=T),,drop=F]
p4 = ggplot(melt(highSeg), aes(Var2, value)) +  ylim(-1,1) + facet_grid(~Var1) + geom_point() + 
  theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA samples', subtitle='patient 512 P', y='5Mb bins')

grid.arrange(p1,p2,p3,p4,ncol=1)

@@

 top='Random BLD or gastric normal', bottom=randNormal)

@@

top='BE', bottom=randBE)

@@

The ploidy adjusted values now need to be mean normalized to match the means of the training data. The training data mean is equivalent to the diploid state and the same is true of the adjusted CN data.
So the mean normalization is simply adding the difference of the means to the SNP bins.
```{r}
training = do.call(bind_rows, purrr::map(list.files('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb', '5e06_tiled_segvals.txt', full.names = T, recursive = T), function(f) {
  read_tsv(f, col_types = cols(.default=col_double(), 'X1'=col_character()))
}))
training.segs = as.matrix(training[,-1]); rownames(training.segs) = training$X1

training = do.call(bind_rows, purrr::map(list.files('~/Data/BarrettsProgressionRisk/Analysis/pcf_perPatient/50kb', 'arms_tiled_segvals.txt', full.names = T, recursive = T), function(f) {
  read_tsv(f, col_types = cols(.default=col_double(), 'X1'=col_character()))
}))
training.arms = as.matrix(training[,-1]); rownames(training.arms) = training$X1
```

```{r raw.adjust, fig.height=16, fig.width=4, echo=T}
median(training.segs)
median(mergedSegs)

median(training.arms)
median(mergedArms)

# difference between the two is 0.98 so...
mergedSegs = mergedSegs + 0.98
mergedArms = mergedArms + 0.98

# Also, for the z-adjust values get a mean+sd from both the training and SNP cohort jointly, due to the difference in the variance
merged = rbind(mergedSegs, training.segs)
merged.mean = apply(merged, 2, mean, na.rm=T)
merged.sd = apply(merged, 2, sd, na.rm=T)

@@

#normal.ids = sample.info %>% filter(grepl('BLD|gastric', Level, ignore.case=T)) %>% dplyr::summarise(rows = paste0(PatientID, '_', EndoID)) %>% pull
#nm.rows = which(rownames(mergedSegs) %in% normal.ids)
```

```{r echo=F,message=F, fig.width=10, fig.height=8}
cx = BarrettsProgressionRisk:::scoreCX(mergedSegs, 1)

arrayDf = BarrettsProgressionRisk:::subtractArms(mergedSegs, mergedArms)
arrayDf = cbind(arrayDf, 'cx'=BarrettsProgressionRisk:::unit.var(cx, mn.cx, sd.cx))

melted = melt(arrayDf) %>% left_join(patient.info %>% dplyr::select(UniqueSampleID, Status), by=c('Var1'='UniqueSampleID'))

train.melt = melt(dysplasia.df) %>% 
  left_join(as_tibble(train.labels) %>% mutate(label = names(train.labels)), by=c('Var1'='label')) %>%
  mutate(value.y = as.factor(value.y))

lims=range(arrayDf, dysplasia.df,na.rm=T)
grid.arrange(
  ggplot(melted, aes(Var2, value,color=Status)) + geom_point() + ylim(lims) +
    geom_point(data=melt(arrayDf[,589:634]), col='red') + labs(title='SNP'),
  ggplot() + geom_point(data=train.melt, aes(Var2, value.x, color=value.y)) + ylim(lims) +
      geom_point(data=melt(dysplasia.df[,589:634]), aes(Var2, value), col='red') + labs(title='sWGS'),
nrow=1)

@@

The SNP and sWGS data is comparable, but it's also clear that the variance is fairly different.  Given our experience with other data this is expected.

@@

For simplicity here I'm not bothering with the 'moderate' category, so the cutoff for high ~ low is a probability of  low<0.5>=high

It's worth noting here that the relative risks are much higher, again I'd attribute this to the variance.

```{r, warning=F, fig.height=8}
prob = predict(fitV, newx=arrayDf, s=l, type='response')
rr = predict(fitV, newx=arrayDf, s=l, type='link')

preds = bind_cols('Samplename'=rownames(prob),'Prob'=prob[,1],'RR'=rr[,1]) %>%
  mutate(Risk = case_when(
    Prob < 0.5 ~ 'Low',
    Prob >= 0.5 ~ 'High' 
))

normals = filter(preds, grepl('BLD|gastric', Samplename, ignore.case=T))

@@

preds = preds %>% left_join(patient.info %>% dplyr::select(PatientID, UniqueSampleID, Pathology, Status, Path.Status), by=c('Samplename'='UniqueSampleID'))

# NP patients who are likely P
np.wgd.pid = qcdata.endo %>% filter(Ploidy_mean > 2.7 & Status == 'NP') %>% ungroup %>% dplyr::select(PatientID) %>% pull
preds = preds %>% dplyr::mutate(Path.Status = as.character(ifelse(PatientID %in% np.wgd.pid, 'P', Path.Status)))

@@

preds %>% group_by(Path.Status, Risk) %>% tally %>% spread(Risk,n)

myPal = rev(RColorBrewer::brewer.pal(11, 'RdYlBu'))
ggplot(bind_rows(preds,normals), aes(RR)) + geom_histogram(aes(fill=..x..), bins=15, show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(y='n Samples', x='Relative Risk', title='Unadjusted relative risk', subtitle='All SNP samples')

ggplot(bind_rows(preds,normals), aes(Prob)) + geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions, all samples model', y='n Samples', x='Unadjusted Probability', subtitle='All SNP samples')
```

The best false positive rate can be assessed from the blood/gastric normals.  So in this case ~17%
`r normals %>% group_by(Risk) %>% tally %>% mutate(ratio = round(n/nrow(normals),2))  %>% kable(caption='Blood & gastric normals') %>% kable_styling(full_width = F)`
